
---
title: "asbur_chapter4"
author: "Ian Bates"
date: "2/24/2022"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(remedy)
remedy_opts$set(name = "Bates")
```

```{r}
qnorm(c(0.975), 0, 1)
```

##### Example 1: Hypothetical Gdf5 gene expression. To illustrate the Ztest we will look at the Gdf5 gene from the Golub et al. (1999) data2. The Gdf5 expression values are contained in row 2058. A quick search through the NCBI site makes it likely that this gene is not directly related to leukemia. Hence, we may hypothesize that the population mean of the ALL expression values equals zero. Accordingly, we test H0 : µ = 0 against Ha : µ 6= 0. For the sake of illustration we shall pretend that the population standard deviation σ is known to be equal to 0.25. The test statistic z-value can be computed as follows:

```{r remedy001}


data(golub, package = "multtest")


gol.fac <- factor(golub.cl,
                  levels = 0:1,
                  labels = c("ALL", "AML")
)

sigma <- 0.25
n <- 27
mu0 <- 0
x <- golub[2058, gol.fac == "ALL"]
z.value <- sqrt(n) * (mean(x) - mu0) / sigma
z.value

2 * pnorm(-abs(z.value), 0, 1)
```




```{r Bates001, tidy=TRUE}

f <- function(x) {
  dnorm(x, 0, 1)
}
x1 <- seq(-4, -1.96, 0.01)
y1 <- dnorm(x1, 0, 1)
x2 <- seq(-1.96, 1.96, 0.01)
y2 <- dnorm(x2, 0, 1)
x3 <- seq(1.96, 4, 0.01)
y3 <- dnorm(x3, 0, 1)
plot(f,
     -4,
     4,
     cex.lab = 1.5,
     xlab = "x",
     ylab = "Normal probability density function f(x)"
)
polygon(c(-4, x1, -1.96), c(0, y1, 0), col = "red")
polygon(c(-1.96, x2, 1.96), c(0, y2, 0), col = "lightblue")
polygon(c(1.96, x3, 4), c(0, y3, 0), col = "red")
arrows(-3, 0.15, -3, 0.03)
text(-3, 0.23, "Rejection")
text(-3, 0.20, "Region")
text(-3, 0.17, expression(alpha / 2))
arrows(3, 0.15, 3, 0.03)
text(3, 0.23, "Rejection")
text(3, 0.20, "Region")
text(3, 0.17, expression(alpha / 2))
text(0, 0.23, "Acceptance")
text(0, 0.20, "Region")

?expression
```




##### Example 2: Confidence interval. Using the data from Example 1, the 95% confidence interval given by Equation 4.1 can be computed as follows:4

```{r}

mean(x) + c(-1, 1) * qnorm(c(0.975), 0, 1) * 0.25 / sqrt(n)


library(TeachingDemos)
z.test(x, mu = 0, sd = 0.25)
?pt

```



```{r}

ci.examp(
  mean.sim = 0,
  sd = 1,
  n = 25,
  reps = 100,
  method = "z",
  lower.conf = 0.025,
  upper.conf = 0.975
)

```
##### 4.1.2 One-sample t-test In almost all research situations, the population standard deviation σ is unknown - and so the above Z-test is usually not applicable. In such cases, t-tests are very useful for testing H0 : µ = µ0 against Ha : µ 6= µ0, where the sample standard deviation s is used in lieu of the population standard deviation σ. The test is based on the test statistic t-value defined by t = √ n(x−µ0)/s. The corresponding p-value is defined by 2·P(Tn−1 ≤ −|t|). As with the Z-test, H0 is not rejected if the p-value is larger than the significance level and H0 is rejected if the p-value is smaller than the significance level. Equivalently, if t falls in the acceptance region (t0.025,n−1, t0.975,n−1), then H0 is not rejected and otherwise it is. For n = 6, the acceptance and rejection regions are illustrated in Figure 4.2. We can create Figure 4.2 using the dt(), polygon(), arrows(), seq(), mtext(), and text() functions:



```{r}
help("~")

f <- function(x) {
  dt(x, 5)
}
x1 <- seq(-4, qt(0.025, 5), 0.01)
y1 <- f(x1)
x2 <- seq(qt(0.025, 5), qt(0.975, 5), 0.01)
y2 <- f(x2)
x3 <- seq(qt(0.975, 5), 4, 0.01)
y3 <- f(x3)
plot(f, -4,
     4,
     xlab = "x",
     ylab = "t-Distribution probability density function f(x)")
polygon(c(-4, x1, qt(0.025, 5)), c(0, y1, 0), col = "red")
polygon(c(qt(0.025, 5), x2, qt(0.975, 5)), c(0, y2, 0), col = "lightblue")
polygon(c(qt(0.975, 5), x3, 4), c(0, y3, 0), col = "red")
arrows(-3, 0.15, -3, 0.03)
text(-3, 0.23, "Rejection")
text(-3, 0.20, "Region")
text(-3, 0.17, expression(alpha / 2))
arrows(3, 0.15, 3, 0.03)
text(3, 0.23, "Rejection")
text(3, 0.20, "Region")
text(3, 0.17, expression(alpha / 2))
text(0, 0.23, "Acceptance")
text(0, 0.20, "Region")
mtext(expression(t[0.025]),
      side = 1,
      at = qt(0.025, 5),
      col = "red")
mtext(expression(t[0.975]),
      side = 1,
      at = qt(0.975, 5),
      col = "red")

```

##### Example 1: Actual Gdf5 gene expression. Let’s test H0 : µ = 0 against Ha : µ 6= 0 for the ALL population mean of the Gdf5 gene expressions. The latter are collected in row 2058 of the golub data. The t-value is computed as follows:
```{r}
x <- golub[2058, gol.fac == "ALL"]; mu0 <- 0; n <- 27
t.value <- sqrt(n) * (mean(x) - mu0) / sd(x)
t.value
```
```{r}
2 * pt(-0.0010, 26)
```

```{r}
c(qt(0.025, n - 1), qt(0.975, n - 1))
```
```{r}
mean(x) + qt(0.025, 26) * sd(x) / sqrt(n)

```


```{r}
t.test(x, mu = 0)

```
##### Example 2: CCND3 gene expression. In Chapter 2, a box-and-whiskers plot revealed that the ALL gene expression values of CCND3 (Cyclin D3) are positive. We can test H0 : µ = 0 against Ha : µ > 0 using the built-in t.test() function. Recall that the corresponding gene expression values for CCND3 are collected in row 1042 of the golub data matrix:
```{r}
ccnd3 <- grep("CCND3",golub.gnames[,2], ignore.case = TRUE)
ccnd3
t.test(golub[ccnd3,gol.fac=="ALL"],mu=0, alternative = c("greater"))
```
#### 4.1.3 Two-sample t-test with unequal variances (Welch’s two-sample t-test)

#### Example 1: CCND3 gene expression. Golub et al. (1999) argue that gene CCND3 (Cyclin D3) plays an important role with respect to discriminating ALL from AML patients. The box plot in Figure 2.5 suggests that the ALL population mean differs from that of AML. The null hypothesis of equal means can be tested by the function t.test() and the appropriate factor and specification var.equal=FALSE:
```{r}
t.test(golub[ccnd3,] ~ gol.fac, var.equal=FALSE)
```
### 4.1.4 Two-sample t-test with equal variances (pooled two-sample t-test)

#### Example 1: CCND3 gene expression. The null hypothesis H0 for gene CCND3 (Cyclin D3) that the mean of the ALL patients differs from that of AML patients can be tested by the two-sample t-test using the specification var.equal=TRUE.

```{r}
t.test(golub[ccnd3,] ~ gol.fac, var.equal=TRUE)
```
#### From the p-value 6.046·10−8, the conclusion is to reject the null hypothesis of equal population means. Note that the p-value is slightly smaller than that of the previous test.


###4.1.5 Paired, two-sample t-test
#### The paired family of tests is adequate for paired difference tests between matched, labeled samples and repeated measurements. For example, if we wish to test the hypothesis that the mean gene expression is the same for the same patients before and after chemotherapy treatment, then we would use a paired, two-sample t-test. In the case of repeated measurements in time, as in the previous example, the paired t-test is also sometimes referred to as the repeated measures t-test.


### 4.1.6 F-test on equal variances